### 2016-04-04 Manuscript/response.tex

%TG: plus response to editors points:
% *Please also confirm in your cover letter whether all the figures are your own or whether permission has been obtained for their use
% *Please confirm whether your manuscript should be available via open access, as we note that this may be a requirement of your funder.
We would like to state that all the figures are ore our own and the the manuscript should be available via open access.
% *If you have any images that can be used to promote your article on social media (should this be accepted) please upload them as a supplementary file
Additionally, we attached a supplementary file named \texttt{article\_promotion.eps} for promoting our article (should it be accepted) on social media. 
% TO DO!

% *Please upload your original figure files as eps, tiff or jpeg files, rather than PDFs
%TG: TO DO (easy)

%%%%%%%%%%%%%%%
% Reviewer 1
%%%%%%%%%%%%%%%

\section{Reviewer 1:}

\begin{enumerate}
\item{\textcolor{blue}{In ESM 1, p. 4, the authors explain their methods of collecting morphological data from the literature. Why select only the first 10 results per search term? How do you know that relevant papers wouldn't come up later in the search? I need evidence to support this cutoff; as it stands, I'm concerned that you're missing some relevant material.}}

As shown in ESM1 Figure 1, the number of OTUs with morphological cladistic data extracted from the Google Scholar searches seems to plateau really quickly (i.e. before the first 100 matches).
Additionally, as Graeme Lloyd mentions, ``older studies tend to be smaller (due to computing limitations) and to commonly reoccur anyway due to data set reuse'' (Reviewer 4, comment 1).
Following these two observations, we think that including the 20 most recent results for each order (and higher taxonomic level) was a good balance between the time spent looking for papers, the number of OTUs extracted and the repeatability of the study.
We are aware however that we might have missed some specific studies (see Reviewer 2 comment 2) but are confident that our protocol gives a good overall picture of the data availability and that the three repositories used before the Google Scholar search (MorphoBank, Graeme Lloyd's and Ross Mounce's) contained most of the available data.
In fact, as mentioned in ESM 1, part 1 ``in the 660 papers [collected from the Google Scholar search], only 50 contained extra living OTUs and only contributed 425 OTUs to our total of 4950 OTUs''.

@@

We replaced the Table 1 in the main text with the Table 1 in the ESM 2 (available data without any number of characters threshold - see reviewer 2 comment 1).
We therefore removed ESM 2 and renamed ESM 3 to ESM 2

\end{enumerate}

%%%%%%%%%%%%%%%
% Reviewer 2 (Graham Slater)
%%%%%%%%%%%%%%%

\section{Reviewer 2 (Graham Slater):}
\begin{enumerate}
\item{\textcolor{blue}{The authors state on page 3 lines 82-90 that they avoided matrices with ``few characters'' due to lower probability of overlapping characters with other matrices and so only selected matrices containing $>$100 characters. First, a lack of overlapping characters doesn’t seem justifiable, at least to my mind, as a reason to exclude a matrix if the ultimate goal is to infer the placement of extinct taxa relative to extant taxa. Antler ramifications, to use the authors' example, are likely to be a very useful character when attempting to resolve relationships among living and extinct artiodactyls, even if they cannot be coded for monotremes. To flip the argument, a matrix of 200 + characters for mysticetes is similarly likely to contain many non-applicable characters from the perspective of talpids. I do understand that the authors are thinking downstream in terms of super matrices for higher level analyses, but smaller matrices are going to be just as relevant for reaching this goal. You don't have much space in the ms I understand, but surely this needs to be addressed?!}}

@@

Again we completely see the reviewer's point and have modified our manuscript as follows.
By removing the threshold (see above) we now included the taxa coded in Wang and Carranza-Casta\~{n}eda 2008 (matrix reference GS2014-WCC) and other matrices excluded in the former analysis increasing the number of Carnivora from 42 species (14\% of total number of species) to 76 (27\%).
However, the two other matrices mentioned by reviewer 2 (Salles 1992 and Bryant et al. 1993) were not included in the new analysis.
This is due to two main reasons: firstly, these two matrices were not available on any of the repositories (either Graeme Lloyd's, Ross Mounce's or MorphoBank's), and secondly, they didn't appear in our additional Google Scholar searches because they were published before 2010.
Following this reviewer's comment as well as reviewer 4's we added a paragraph in the discussion to clearly state that we are aware of data sampling biases:

``It is worth noting, however, that our analysis did not include all the matrices containing anatomical characters ever published.
In fact, our data collection procedure focused on including studies that provided matrices easily accessible, i.e. we did specifically not include any matrices that were only available in paper format (e.g. printed in books), non-reusable format (e.g. an image of the matrix)  or/and matrices available only upon request (e.g. by emailing the authors).
The difficulty of obtaining such matrices makes them less likely to be included in Total Evidence analyses, especially where the number of taxa to include in the phylogeny is high.
'' p.5 lines 161-168.

\item{\textcolor{blue}{I finally am a little skeptical of the sampling done. Looking at Figure 1B, again the clade I'm most familiar with, representation for carnivorans appears very poor. Notwithstanding the issues discussed above regarding matrices with $<$100 characters, there are definitely matrices out there that fulfill the authors' requirements that are not represented. For example, a 349 character matrix with near complete sampling at the species is available for Viverridae, Prionodontidae, and Eupleridae in Gaubert et al (2005 Syst Biol.). Going through each clade in the supplement seems beyond the scope of reviewing duties, but I am concerned that if this one matrix can be missed others may have been missed too. I think a reexamination of the literature may be warranted using taxon specific search terms.}}

We dealt with this comment by removing the matrix size threshold and adding a paragraph to the discussion explaining that some matrices might not be included in this study (see reviewer's 2 comments 1 and 2).
Concerning the Gaubert et al. 2005's study specifically, this study unfortunately did not contain any easily available matrix.
In fact, the matrix mentioned as Appendix 2 in the paper is not present in the article and neither of the TreeBase accession numbers (``S1255'', ``M2193'') nor the provided URL (``www.systematicbiology.org'') links to the matrix.

@@

This is a great suggestion, thanks!
We added this analysis to the ESM 2 (Supplementary results - 1) showing, as suggested, that mammalian morphological phylogeneticists have focused their energies on resolving the higher level relationships among mammalian lineages.

\end{enumerate}

%%%%%%%%%%%%%%%
% Reviewer 3 (Peter Wagner)
%%%%%%%%%%%%%%%

\section{Reviewer 3 (Peter Wagner):}
\begin{enumerate}
\item{\textcolor{blue}{My biggest criticism is a semantic one: Morphological data $\neq$ cladistic data.}}
We agree with this reviewers semantic point and changed the first mention of ``cladistic'' data to ``discrete morphological/anatomical data'' (p.2 line 38) and then to ``anatomical characters'' throughout the manuscript.

@@

\item{\textcolor{blue}{Lines 70-71: ``We downloaded all cladistic matrices containing any living and/or fossil mammal  taxa from three major public databases:''
As a side note and for future reference for the authors, a large number of phylogenetic data sets for mammals (and other metazoans) also are available at the Paleobiology Database.
(These were formerly at my website at the Field Museum, but that site obviously disappeared when I moved to the Smithsonian.)
Because of the nature of the projects for which I was using these datasets, most of them are species-level, but some are genus-level, too.}}

We thank the reviewer for this mention and we will be sure to use this database for future studies.

\item{\textcolor{blue}{Lines 164-166: ``For example, a Carnivora fossil will be unable to branch in the Herpestidae, and will have more chance to randomly branch within Canidae (Figure 1B).'' 
This seems to be an awkward way to phrase this.
Moreover, this seems like an extension of the long-branch problem.
Basically, the issue really is that we do not have coded skeletal characters for extant herpestids.
Thus, it is not possible for coded skeletal characters from fossil herpestids to connect those fossil herpestids to the extant herpestids.
This is the opposite of Felsenstein's long branch: insofar as skeletal characters are concerned, extant herpestids have zero-length branches.
That in turn means that the fossil herpestids are going to be ``attracted'' to non-zero length branches elsewhere, with long branches in related groups such as canids being particularly good candidates because there is a higher chance of homoplasies between canids and herpesitds on the longer branches.
So, it will not be random: they probably will latch themselves onto the longest reasonably branch among the closest relatives for which anatomical data are present.}}

We agree with this comment and changed this to:
``For example, a Carnivora fossil will be unable to be placed in the Herpestidae clade because they have no coded anatomical characters available.
Instead the fossil will have a high probability of being placed on a branch that contains many anatomical characters such as within the Canidae or Ursidae (Figure 1B). 
This is analogous to the problem of long-branch attraction/short branch repulsion, as one can think of herpestids as having zero-length branches for anatomical characters, and canids and ursids having long branches and thus ``attracting'' fossil placements.'' p.5 lines 154-160.

@@

%%%%%%%%%%%%%%%
% Reviewer 4 (Graeme Lloyd)
%%%%%%%%%%%%%%%

\section{Reviewer 4 (Graeme Lloyd):}
\begin{enumerate}
\item{\textcolor{blue}{Reviewer 1's most substantial complaints (the year and page number cut offs for google scholar searches) seem at least partially valid.
However, in my experience older studies tend to be smaller (due to computing limitations) and to commonly reoccur anyway due to data set reuse.
The authors also already state that subsequent search results tended to yield no substantial additional data, and by the reviewer's own admission the assembled data seem to look correct.
Thus a recollecting of the data doesn't seem necessary (but see further comments below).}}

We developed the description of our data collection protocol in the ESM 1 (see Reviewer 1 comment 2) and underlined the caveats in the discussion (see Reviewer 2 comment 2).

\item{\textcolor{blue}{Reviewer 2 (Graham Slater)'s complaint are - in my view - the most substantial.
I agree that the $>$100 characters cut off is not as strongly justified as it might be.
However, as the authors have run their protocol without this threshold this is easily fixed by at least briefly mentioning how this affects their results in the main text (not relegating it to the ESM where no actual discussion is provided).
From my reading of the two tables only 11 of the orders seem to have below 25\% coverage at the species-level (not 22 as the authors state in the main text!) when the threshold is used, and this improves slightly to 10 for when the threshold is removed.
As this difference isn't dramatic it seems worth mentioning.
Reviewer 2 also notes that some important studies may be missed.
This seems valid (I am aware of the biases in Ross Mounce's data collection protocol, and obviously my own) and I am sure there are many more data sets that could be sampled.
In particular I am aware of a number of studies that can only be found in books and hence are not easily accessed even if they occur in a google scholar search.
I do not expect the authors to more thoroughly search the literature for all possible cladistic analyses of mammals, however, as this is a major task.
(I should know, it has been a long term project of mine for years.)
Nevertheless, it does deserve mention in the discussion as a potential bias in the results.
I don't think this undermines the study, but rather should be viewed as both a reason for optimism and an acknowledgement of a separate problem (i.e., access to cladistic data) which multiple publications already cover (e.g., see Ross Mounce's work).}}

@@

We added the data coverage among all mammals in the Table 1 (``Mammalia (Class)'') and added it to the methods, results and discussion.
We also added the Reviewer 2's extra analysis suggestion to the ESM 2 part 1.

\end{enumerate}

\subsection{Reviewer 4 specific comments:}
\begin{enumerate}
\item{\textcolor{blue}{Lines 72-74 - Web addresses should be consistent in dropping of ``http://www.''.}}
We removed the ``http://www.'' part of the URLs.

\item{\textcolor{blue}{Line 104 - ``NTI 15'' should probably be ``NTI [15]''.}}
We fixed this typo.

@@

\item{\textcolor{blue}{ESM 2 - ``The following section contains supplementary results to the main body: the available data structure using the NTI and the PD metric'' Abbreviations should be introduced when first used, not later as they are here.}}
We changed this to ``using the Net Relatedness Index (NRI) and the Nearest Taxon Index (NTI) metrics''.

\end{enumerate}

We hope we have responded to all these comments appropriately. Please let us know if you require any further information,\\
\bigskip

Thomas Guillerme (t.guillerme@imperial.ac.uk),\\
Natalie Cooper

### 2016-04-04 Manuscript/Missing_mammals-draft.tex
color testings
Cleaned up manuscript file

\noindent{\Large \bf Morphological data is lacking for living mammals}
%Or: Missing cladistic data in living mammals - but I prefer the first one emphasizing on the availability of it. The data is not missing per se, it's just not coded!
% NC: sometimes the better title is one that answers the question rather than giving the topic. e.g., "Morphological data is lacking for living species" or similar
% TG: maybe worth considering the special issue as well? "Putting fossils in tree" so maybe something like "There is not enough morphological data for living mammals to branch fossils in the trees."
% TG: or some poetic... "Living and dead mammals: morphological data is lacking in living for combined analysis"
\bigskip

\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^*$ and Natalie Cooper$^1$$^,$$^2$}\\
\noindent {\small \it 
$^1$School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland.\\
$^2$Department of Life Sciences, Natural History Museum, Cromwell Road, London, SW7 5BD, UK.}\\
\end{center}
\medskip
\noindent{*\bf Corresponding author.} \textit{Zoology Building, Trinity College Dublin, Dublin 2, Ireland; E-mail: guillert@tcd.ie; Fax: +353 1 6778094; Tel: +353 1 896 2571.}\\  
\vspace{1in}

%Line numbering REMOVE FOR WORDCOUNT
%\modulolinenumbers[1]
%\linenumbers

%---------------------------------------------
%
%       ABSTRACT
%
%---------------------------------------------
% NC: Time to sort this out!
\newpage
\begin{abstract} % 200 words, all keywords written in abstract if possible. TG: That's 195 words for now.
Combining living and fossil in the same analysis data is crucial for studying changes in global biodiversity through time.
One method allowing to combine this data is the Total Evidence method that uses both molecular data for living species and morphological data for both living and fossil species.
With this method, a good overlap of morphological data between living and fossil taxa is crucial for accurately inferring the phylogenies’ topology.

Since the advent of DNA, molecular data has become easily and widely available.
However, despite two centuries of morphological studies, scientists using and generating such data mainly focus on palaeontological data.
Therefore, there is a gap in our knowledge of neontological morphological data even in well studied groups such as mammals.

In this study, we quantify the morphological data available for living mammal taxa.
We then analyse the structure of the available data by testing if it is clustered or evenly spread across the phylogeny.
We found that 78\% of mammalian orders have less than 25\% data available at the species level.
However, we found that the available is often randomly distributed among these orders apart from six of them where the data is clustered
\end{abstract}

%\noindent Key words: Total Evidence method, data structure, phylogenetic, fossil, topology\\ Excluded from word count.

@@

%---------------------------------------------
%
%       INTRODUCTION
%
%---------------------------------------------
\newpage 
\section{Introduction}
There is an increasing consensus among evolutionary biologists that studying both living and fossil taxa is essential for fully understanding macroevolutionary patterns and processes \cite{slaterunifying2013,fritzdiversity2013,Wood01032013}.
For example, including both living and fossil taxa in evolutionary studies can improve the accuracy of timing diversification events (e.g. \cite{ronquista2012}, our understanding of relationships among lineages (e.g. \cite{beckancient2014}, and our ability to infer biogeographical patterns through time (e.g. \cite{Meseguer01032015}.
To perform such analyses it is necessary to combine living and fossil taxa in phylogenetic trees.
One increasingly popular method, the Total Evidence method \cite{eernissetaxonomic1993,ronquista2012}, combines molecular data from living taxa and morphological data from both living and fossil taxa in a supermatrix (e.g. \cite{pyrondivergence2011,ronquista2012,schragocombining2013,slaterunifying2013,beckancient2014,Meseguer01032015}, producing a phylogeny with living and fossil taxa at the tips. 
These phylogenies can be dated using methods such as tip-dating \cite{ronquista2012,Wood01032013} and incorporated into macroevolutionary studies (e.g. \cite{ronquista2012,Wood01032013,slaterphylogenetic2013}.

A downside of the Total Evidence method is that it requires a lot of data.
One must collect molecular data for living taxa and morphological data for both living and fossil taxa; two types of data that require fairly different technical skills (e.g. molecular sequencing \textit{vs.} anatomical description).
Additionally, large chunks of this data can be difficult, or even impossible, to collect for every taxon present in the analysis.
For example, fossils very rarely have molecular data and incomplete fossil preservation (e.g. soft \textit{vs.} hard tissues) may restrict the amount of morphological data available \cite{sansomfossilization2013}.
Additionally, since the molecular phylogenetics revolution, it has become less common to collect morphological characters for living taxa when molecular data is available (e.g. in \cite{slaterphylogenetic2013}, only 13\% of the 169 living taxa have coded morphological data).
Unfortunately this missing data can lead to errors in phylogenetic inference; in fact, simulations show that the ability of the Total Evidence method to recover the correct phylogenetic topology decreases when there is a low overlap between morphological data in the living and fossil taxa \cite{GuillermeCooper}, regardless the overall amount of morphological data available for the fossils (or the amount of molecular data available for the living species).
The effect of missing data on topology is greatest when living taxa have few morphological data.
This is because (1) a fossil cannot branch in the correct clade if there is no overlapping morphological data in the clade; and (2) a fossil has a higher probability of branching within a clade with more morphological data available for living taxa, regardless of whether this is the correct clade or not \cite{GuillermeCooper}.

The issues above highlight that it is crucial to have sufficient morphological data for living taxa in a clade before using a Total Evidence approach.
However, it is unclear how much morphological data for living taxa is actually available (i.e. already coded from museum specimens and deposited in phylogenetic matrices accessible online), and how this data is distributed across clades.
Intuitively, most people assume this kind of data has already been collected, but empirical data suggest otherwise (e.g. in \cite{ronquista2012,slaterphylogenetic2013,beckancient2014}).
To investigate this further, we assess the amount of available morphological data for living mammals to determine whether sufficient data exists to build reliable Total Evidence phylogenies in this group.
We collected cladistic data (i.e. discrete morphological characters used in phylogenetics) from 286 phylogenetic matrices available online and measured the proportion of cladistic data available for each mammalian order.
%Additionally, if the available data is randomly distributed across clades, the topological errors should be less extreme than if the data is biased towards particulars groups.
Additionally, because missing morphological data in living species can influence tree topology as described above, %NC: Might still need to fix this sentence. "Topological caveats above" was too unclear.
we determined whether the available cladistic data was phylogenetically overdispersed or clustered in the mammalian orders where data was missing. 
We find that available morphological data for living mammals is scarce but generally randomly distributed across phylogenies. 
We recommend that efforts be made to collect and share more cladistic data for living species to improve the accuracy of Total Evidence phylogenies.

% NC: If there are word count problems we can ditch the last two sentences.
%---------------------------------------------
%
%       METHODS
%
%---------------------------------------------
\section{Materials and Methods}
\subsection{Data collection and standardisation}
We downloaded all cladistic matrices containing any living and/or fossil mammal taxa from three major public databases (accessed 10th of June 2015): Morphobank (\url{http://www.morphobank.org/} \cite{morphobank}), Graeme Lloyd's website (\url{graemetlloyd.com/matrmamm.html}) and Ross Mounce's GitHub repository (\url{https://github.com/rossmounce/cladistic-data}).
We also performed a systematic Google Scholar search (accessed 11th of June 2015) for matrices that were not uploaded to these databases (see Supplementary Materials for a detailed description of the search procedure).
In total, we downloaded 286 matrices containing a total of 11010 operational taxonomic units (OTUs) of which 5228 were unique. % Number of unique OTUs (including fossils)
In this study, we refer to OTUs rather than species since the entries in the downloaded matrices were not standardised and ranged from specific individual specimen names (i.e. the name of a collection item) to the family-level.
Where possible, we considered OTUs at their lowest valid taxonomic level (i.e. species) but some OTUs were only valid at a higher taxonomic level (e.g. genus or family).
Therefore for some orders, we sampled more genera than species (Table \ref{Table_morpho_taxa_proportion}).

To select the lowest valid taxonomic level for each OTU, we standardised their taxonomy by correcting species names so they matched standard taxonomic nomenclature (e.g., \textit{H. sapiens} was transformed to \textit{Homo sapiens}).
We designated as ``living'' all OTUs that were either present in the phylogeny of \cite{BinindaEmonds} or the taxonomy of \cite{wilson2005mammal}, and designated as ``fossil'' all OTUs that were present in the Paleobiology database (\url{https://paleobiodb.org/}).
For OTUs that did not appear in these three sources, we first decomposed the name (i.e. \textit{Homo sapiens} became \textit{Homo} and \textit{sapiens}) and tried to match the first element with a higher taxonomic level (family, genus etc.).
Any OTUs that still had no matches in the sources above were designated as non-applicable (NA; see Supplementary Material for more details).

The number of characters in each matrix ranged from 6 to 4541.
Matrices with few characters are problematic when comparing available data among matrices because (1) they have less chance of having characters that overlap with those of other matrices \cite{wagner2000} and (2) they are more likely to contain a higher proportion of specific characters that are not-applicable across large clades (e.g. ``antler ramifications'' is a character that is only applicable to Cervidae not all mammals \cite{Brazeau2011}).
Therefore we selected only matrices containing \textgreater 100 characters for each OTU.
This threshold was chosen to correspond with the number of characters used in \cite{GuillermeCooper} and \cite{harrisonamong-character2014}.
Note that results of analyses with no character threshold are available in Supplementary Material. 
After removing all matrices with \textless 100 characters, we retained 1074 unique living mammal OTUs from 126 matrices for our analyses. % 1601 unique living OTUs for 286 matrices (no threshold)

\subsection{Data availability and distribution}
To assess the availability of cladistic data for each mammalian order, we calculated the percentage of OTUs with cladistic data at three different taxonomic levels: family, genus and species.
%We used these different taxonomic levels because some clades are well covered at the family- or genus-level, but poorly covered at the species-level. % NC: Probably obvious?
We consider orders with \textless 25\% of living taxa with cladistic data as having poor data coverage (``low'' coverage), and orders with \textgreater 75\% of living taxa with cladistic data as having good data coverage (hereafter ``high'' coverage).

For orders with \textless 100\% cladistic data coverage at any taxonomic level, we investigated whether the available cladistic data was (i) randomly distributed, (ii) overdispersed or (iii) clustered, with respect to phylogeny, using two metrics from community phylogenetics: the Nearest Taxon Index (NTI; \cite{webb2002phylogenies}) and the Net Relatedness Index (NRI; \cite{webb2002phylogenies}). 
NTI is most sensitive to clustering or overdispersion near the tips, whereas NRI is more sensitive to clustering or overdispersion across the whole phylogeny \cite{Cooper2008}. 
Both metrics were calculated using the \texttt{picante} package in R \cite{picante,R}.

NTI \cite{webb2002phylogenies} is based on mean nearest neighbour distance (MNND) and is calculated as follows:
  \begin{equation}
    NTI=-\left(\frac{\overline{MNND}_{obs}-\overline{MNND}_{n}}{\sigma(MNND_{n})}\right)
  \end{equation}
where $\overline{MNND}_{obs}$ is the observed mean distance between each of $n$ taxa with cladistic data and its nearest neighbour with cladistic data in the phylogeny, 
$\overline{MNND}_{n}$ is the mean of 1000 mean MNND between $n$ randomly drawn taxa, and $\sigma(MNND_{n})$ is the standard deviation of these 1000 random MNND values.
NRI is similar but is based on mean phylogenetic distance (MPD) as follows:
  \begin{equation}
    NRI=-\left(\frac{\overline{MPD}_{obs}-\overline{MPD}_{n}}{\sigma(MPD_{n})}\right)
  \end{equation}
where $\overline{MPD}_{obs}$ is the observed mean phylogenetic distance of the tree containing only the $n$ taxa with cladistic data, $\overline{MPD}_{n}$ is the expected random MPD for $n$ taxa estimated by calculating the MPD from $n$ taxa randomly drawn from the phylogeny and repeated 1000 times, and $\sigma(MPD_{n})$ is the standard deviation of the 1000 random MPD values. % NC: This can be shortened a bit but it's only gonna save you 50 words max.
Negative NTI and NRI values show that the focal taxa are more overdispersed across the phylogeny than expected by chance, and positive values reflect significant clustering.

We calculated NTI and NRI values for each mammalian order separately, at each different taxonomic level. 
For each analysis our focal taxa were those with available cladistic data at that taxonomic level and the phylogeny was the phylogeny of the order pruned from \cite{BinindaEmonds}.

@@

\section{Results}
Across the 126 cladistic matrices we extracted, 22 out of 28 mammalian orders have low coverage (\textless 25\% of species with cladistic data) and six have high coverage (\textgreater 75\% of species with cladistic data) at the species-level.
At the genus-level, three orders have low coverage and 12 have high coverage, and at the family-level, no orders have low coverage and 23 have high coverage (Table \ref{Table_morpho_taxa_proportion}).

\input{Table_results}
% NC: Can probably cut words from the table legend too

Among the mammalian orders containing OTUs with no available cladistic data, only six orders had significantly clustered data (Carnivora, Cetartiodactyla, Chiroptera and Soricomorpha at both species- and genus-level and Afrosoricida and Pholidota at the species-level only) and no order had significantly overdispersed data at any taxonomic level (Table~\ref{Table_data_structure}).

Two contrasting results are shown in Figure \ref{Figure_example_coverage} with randomly distributed OTUs with cladistic data in Primates (Figure \ref{Figure_example_coverage}A) and phylogenetically clustered OTUs with cladistic data in Carnivora (mainly Canidae; Figure \ref{Figure_example_coverage}B).

\begin{figure}[!htbp]
\centering
    \includegraphics[width=1\textwidth]{example_coverage.pdf}
\caption{Phylogenetic distribution of species with available cladistic data across two mammalian orders (A: Primates; B: Carnivora).
Edges are colored in grey when no cladistic data is available for a species and in blue when data is available.}
\label{Figure_example_coverage}
\end{figure}

@@

\section{Discussion}
Our results show that although phylogenetic relationships among living mammals are well-resolved (e.g. \cite{BinindaEmonds,meredithimpacts2011}) , most of the data used to build these phylogenies is molecular, and very little cladistic data is available for living mammals compared to fossil mammals (e.g. \cite{O'Leary08022013,ni2013oldest}).
This has implications for building Total Evidence phylogenies containing both living and fossil mammals, as without sufficient cladistic data for living species, fossil placements in these trees are very uncertain \cite{GuillermeCooper}.
Cladistic data coverage in living mammals varies across taxonomic levels and in its phylogenetic distribution.
Higher taxonomic levels are always better sampled than lower ones and within these taxonomic levels, the available data is mostly randomly distributed across the phylogeny, apart from in six orders).

>>>>

\noindent{\Large \bf Assessment of available anatomical characters for linking living mammals to fossil taxa in phylogenetic analyses}

\bigskip

\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^*$ and Natalie Cooper$^1$$^,$$^2$}\\
\noindent {\small \it 
$^1$School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland.\\
$^2$Department of Life Sciences, Natural History Museum, Cromwell Road, London, SW7 5BD, UK.}\\
\end{center}
\medskip
\noindent{*\bf Corresponding author.} \textit{t.guillerme@imperial.ac.uk}\\ 
\vspace{1in}

%Line numbering
\modulolinenumbers[1]
\linenumbers

%---------------------------------------------
%
%       ABSTRACT
%
%---------------------------------------------
\newpage
\begin{abstract}
\section{abstract}

Analyses of living and fossil taxa are crucial for understanding biodiversity through time.
The Total Evidence method allows living and fossil taxa to be combined in phylogenies, using molecular data for living taxa and morphological data for living and fossil taxa.
With this method, substantial overlap of coded anatomical characters among living and fossil taxa is vital for accurately inferring topology.
However, although molecular data for living species are widely available, scientists generating morphological data mainly focus on fossils.
Therefore, there are fewer coded anatomical characters in living taxa, even in well-studied groups like mammals.

We investigated the number of coded anatomical characters available in phylogenetic matrices for living mammals and how these were phylogenetically distributed across orders.
11 of 28 mammalian orders have \textless 25\% species with available characters; this has implications for the accurate placement of fossils, although the issue is less pronounced at higher taxonomic levels. 
In most orders, species with available characters are randomly distributed across the phylogeny, which may reduce the impact of the problem.
We suggest that increased morphological data collection efforts for living taxa are needed to produce accurate Total Evidence phylogenies. 
\end{abstract}

\noindent Key words: Total Evidence method, phylogenetic clustering, discrete morphological matrix, extinct, topology.\\

@@

%---------------------------------------------
%
%       INTRODUCTION
%
%---------------------------------------------
\newpage 
\section{Introduction}
There is an increasing consensus among biologists that studying both living and fossil taxa is essential for fully understanding macroevolutionary patterns and processes \cite{slaterunifying2013,fritzdiversity2013}.
To perform such analyses it is necessary to combine living and fossil taxa in phylogenetic trees.
One increasingly popular method, the Total Evidence method \cite{ronquista2012}, combines molecular data from living taxa and morphological data from both living and fossil taxa in a supermatrix that can then be used with the tip-dating method (e.g. \cite{pyrondivergence2011,ronquista2012,schragocombining2013,slaterunifying2013,beckancient2014}), producing a chronogram with living and fossil taxa at the tips.
A downside of this method is that it requires molecular data for living taxa and discrete morphological/anatomical data shared among both living and fossil taxa (i.e. hard tissue characters such as skeletal features).
Sections of these data can be difficult, or impossible, to collect for every taxon in the analysis.
For example, fossils rarely have molecular data and incomplete fossil preservation may reduce the number of anatomical characters available.
Additionally, it has become less common to collect anatomical characters for living taxa when molecular data are available (e.g. in \cite{slaterphylogenetic2013}, only 13\% of living taxa have coded anatomical characters).
Unfortunately this missing data can lead to errors in phylogenetic inference.
We might expect the Total Evidence method to perform poorly when there is little overlap between coded anatomical characters in living and fossil taxa, because fossil taxa cannot be correctly placed within a clade of living species with no coded characters.
Furthermore, simulations show that fossils are more likely to be placed in clades for which more characters have been coded, regardless of whether this is the correct clade \cite{GuillermeCooper}.

The issues above highlight that it is crucial to have sufficient coded anatomical characters available for living taxa in a clade before using the Total Evidence approach.
However, it is unclear how many coded anatomical characters are actually available for living taxa, i.e. already coded from museum specimens and deposited in phylogenetic matrices accessible online, and how these data are distributed across clades.
Intuitively, most people assume these data have already been collected, but empirical analyses suggest otherwise (e.g. in \cite{ronquista2012,slaterphylogenetic2013,beckancient2014}).
To investigate this further, we assess the number of available coded anatomical characters for living mammals to determine whether enough data exist to build reliable Total Evidence phylogenies.
We also determine whether the characters are phylogenetically overdispersed or clustered across mammalian orders.

%---------------------------------------------
%
%       METHODS
%
%---------------------------------------------
\section{Materials and Methods}
\subsection{Data collection and standardisation}
We downloaded all discrete morphological matrices containing any living and/or fossil mammal taxa from three major public databases: MorphoBank (\url{morphobank.org} \cite{morphobank}), Graeme Lloyd's website (\url{graemetlloyd.com/matrmamm.html}) and Ross Mounce's GitHub repository (\url{github.com/rossmounce/cladistic-data}).
We also performed a systematic Google Scholar search for matrices that were not uploaded to these databases (see Electronic Supplementary Material 1 (ESM1) for details).
In total, we downloaded 286 matrices containing 5228 unique operational taxonomic units (OTUs). 
We used OTUs rather than species because entries in the matrices ranged from species to families. 
We standardised the taxonomy as described in the ESM1 and excluded OTUs that were not present in the phylogeny of \cite{BinindaEmonds} or the taxonomy of \cite{wilson2005mammal} to remove fossil species.
This resulted in 1601 unique OTUs from 286 matrices.

\subsection{Data availability and distribution}
To assess the availability of coded anatomical characters for each mammalian order and across mammals, we calculated the percentage of OTUs with coded anatomical characters at three different taxonomic levels: family, genus and species.
We do not distinguish between soft and hard characters, but the majority of matrices contain at least some hard tissue characters. 
We consider orders with \textless 25\% of living taxa with available anatomical characters as having low data coverage, and orders with \textgreater 75\% of living taxa with available anatomical characters as having high data coverage.

For each order and for all mammals, we investigated whether the available coded anatomical characters were (i) randomly distributed, (ii) overdispersed or (iii) clustered, with respect to phylogeny, using two metrics from community phylogenetics: the Nearest Taxon Index (NTI; \cite{webb2002phylogenies}) and the Net Relatedness Index (NRI; \cite{webb2002phylogenies}). 
NTI is most sensitive to clustering or overdispersion near the tips, whereas NRI is more sensitive to them across the whole phylogeny \cite{Cooper2008}. 
Both metrics were calculated using the \texttt{picante} package in R \cite{picante,R}.

NTI \cite{webb2002phylogenies} is based on mean nearest neighbour distance ($MNND$) and is calculated as follows:
  \begin{equation}
    \text{NTI}=-\left(\frac{\overline{MNND}_{obs}-\overline{MNND}_{n}}{\sigma(MNND_{n})}\right)
  \end{equation}
where $\overline{MNND}_{obs}$ is the observed mean sum of the branch lengths between each of $n$ taxa with available coded anatomical characters and its nearest neighbour with available coded anatomical characters in the phylogeny, 
$\overline{MNND}_{n}$ is the mean of 1000 $MNND$ between $n$ randomly drawn taxa, and $\sigma(MNND_{n})$ is the standard deviation of these 1000 random $MNND$ values.
NRI is calculated in the same way, but using the mean phylogenetic distance ($MPD$):
  \begin{equation}
    \text{NRI}=-\left(\frac{\overline{MPD}_{obs}-\overline{MPD}_{n}}{\sigma(MPD_{n})}\right)
  \end{equation}
where $\overline{MPD}_{obs}$ is the observed mean phylogenetic branch length of the tree containing only the $n$ taxa with available coded anatomical characters.
Negative NTI and NRI values show that the focal taxa are more overdispersed across the phylogeny than expected by chance, and positive values reflect clustering.

We calculated NTI and NRI values for all mammals or each mammalian order separately, at each different taxonomic level. 
For each analysis our focal taxa were those with available coded anatomical characters at that taxonomic level and the phylogeny was the order pruned from \cite{BinindaEmonds}.

@@

\section{Results}
Across mammals, species coverage was low (\textless 25\% species with available coded anatomical characters) but family coverage was high (\textgreater 75\% families with available coded anatomical characters).
For each order, 11 out of 28 had low coverage and seven had high coverage at the species-level.
At the genus-level, one order had low coverage and 15 had high coverage, and at the family-level, no orders had low coverage and 25 had high coverage (Table\ref{Table_results}).

\input{Table_results}

Across mammals, taxa with available coded anatomical characters were significantly clustered using NTI at the species- and genus-level.
For each order, only seven showed significant clustering (Cetartiodactyla, Cingulata, Pilosa and Rodentia at the species-level and Carnivora, Chiroptera and Soricomorpha at both species- and genus-level) and none showed significant overdispersion (Table~\ref{Table_results}).

Figure \ref{Figure_example_coverage} shows randomly distributed OTUs with available coded anatomical characters in Primates (Figure \ref{Figure_example_coverage}A) and phylogenetically clustered OTUs with available coded anatomical characters in Carnivora (mainly Canidae and Urisdae but no Herpestidae; Figure \ref{Figure_example_coverage}B).

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{figure1.pdf}
\caption{Phylogenetic distribution of species with available coded anatomical characters across two orders (A: Primates; B: Carnivora).
Blue branches indicate species with available coded anatomical characters.}
\label{Figure_example_coverage}
\end{figure}

@@

\section{Discussion}
Our results show that although phylogenetic relationships among living mammals are well-resolved (e.g. \cite{BinindaEmonds,meredithimpacts2011}), most of the data used to build these phylogenies is molecular, and few coded anatomical characters are available for living mammals compared to fossils (e.g. \cite{O'Leary08022013,ni2013oldest}).
This has implications for building Total Evidence phylogenies, as without sufficient overlapping anatomical characters for living and fossil species, fossil placements in these trees may be unreliable \cite{GuillermeCooper}.
